---
title: "Analyze & Share"
author: "kk"
date: '2022-05-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyze and share
```{r read}
## read merged file from summary csv created in last step
getwd()
library(readr)
library(tidyverse)
df <- read_csv("output-data/allbikedata-clean_onlyvalidridelength.csv")
```

###__Some descriptive statistics:__
```{r descriptive}
## some descriptive stats
library(skimr)
skim(df)
glimpse(df)
```

* __Average ride length and ride count for members and casual riders:__
```{r average ride length and ride count}
# calculate average ride length for members and casual riders
df %>% group_by(member_casual) %>% summarise(mean_length=mean(ride_length_num), 
                                             sd_length=sd(ride_length_num))
print(mean(df$ride_length))
tapply(df$ride_length, df$member_casual, summary)
colnames(df)

# count number of casual and member rides
df %>% group_by(member_casual) %>% summarise(count = n())
```

* __Bike type and weekdays for members and casual riders:__
```{r bike type and weekdays}
# count bike type by members and casual riders
df %>% group_by(member_casual, rideable_type) %>% summarise(count = n())

# count weekdays by members and casual riders
df %>% group_by(member_casual, day_of_week) %>% summarise(count = n())
```

## Visualizations
```{r count member types}
library(ggplot2)
# bar plot of member and causal count
p <- ggplot(data=df, aes(x=member_casual)) +
  geom_bar(fill = "#00abff") +
  geom_text(stat="count", aes(label=scales::comma(..count..)), vjust = -0.5)
## format labels to show actual numbers and not abbreviated
require(scales)
p + scale_y_continuous(labels = comma)
```

* __Interpretation:__ There are 800,000 more rides from people in the 'annual member' category than in the 'casual rides' category. We don't know here if there are more users in one or the other category in total. 

```{r pie charts for member types}
# pie charts for bike type by member group
## make dfs with only member or casual type each
df_m <- subset(df, member_casual == "member")
df_c <- subset(df, member_casual == "casual")

## pie chart for members
### make summary df with count as value because the pie chart does not work for me with the actual count
df_m_sum <- df_m %>% group_by(rideable_type) %>% summarise(n = n())
print(df_m_sum)
ggplot(df_m_sum, aes(x="",y=n, fill=rideable_type)) +
  geom_bar(stat="identity", width = 1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
  theme(legend.position = "none", plot.title = element_text(hjust =0.5)) +
  geom_text(aes(y= n, label=scales::comma(n)), color = "white", size=4, vjust=-6) +
  geom_text(aes(y= n, label=rideable_type), color = "white", size=4, vjust=-8) +
  ggtitle("Rideable types of annual members")
## pie chart for casual riders
df_c_sum <- df_c %>% group_by(rideable_type) %>% summarise(n = n())
print(df_c_sum)
ggplot(df_c_sum, aes(x="",y=n, fill=rideable_type)) +
  geom_bar(stat="identity", width = 1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
  theme(legend.position = "none", plot.title = element_text(hjust =0.5)) +
  geom_text(aes(y= n, label=scales::comma(n)), position = position_stack(vjust = 0.5), color = "white", size=4) +
  geom_text(aes(y= n, label=rideable_type), position = position_stack(vjust = 0.5), color = "white", size=4, vjust=2.5) +
  ggtitle("Rideable types of casual riders")
```

* __Interpretation:__ Annual members only use two categories of bikes, casual members use all three. Maybe docked bikes are only available for casual members, or all annual members are automatically provided with a regular or electric bike. Classic bikes are by far the most popular version for both member types. Docked bikes are the least used version. 

```{r histogram}
# histogram for ride lengths
## for members
## does not look like a helpful plot. 
#ggplot(df_m, aes(x=ride_length_num)) +
#  geom_histogram(bins = 6) +
#  scale_x_continuous(labels = comma) +
#  scale_y_continuous(labels = comma)
print(mean(df_m$ride_length_num))
## cut off everything under 10th and over 90th quantile
quantile(df_m$ride_length_num, prob=c(.10,.5,.90))
df_m_small <- subset(df_m, (ride_length_num > 196 & ride_length_num < 1555))
mean(df_m_small$ride_length_num)
median(df_m_small$ride_length_num)
#skim(df_m_small)
df_m_small$ride_length_min <- df_m_small$ride_length_num/60
#skim(df_m_small)

ggplot(df_m_small, aes(ride_length_min)) +
  geom_histogram(binwidth = 0.5, col = "black", fill = "cornflowerblue") +
  labs(title="Frequency of ride lengths for annual members",
       subtitle="From 10th to 90th quantile",
       x="ride length in minutes")
## for casual riders
#ggplot(df_c, aes(x=ride_length_num)) +
#  geom_histogram(bins = 6) +
#  scale_x_continuous(labels = comma) +
#  scale_y_continuous(labels = comma)
print(mean(df_c$ride_length_num))

quantile(df_c$ride_length_num, prob=c(.10,.5,.90))
df_c_small <- subset(df_c, (ride_length_num > 317 & ride_length_num < 3217))
mean(df_c_small$ride_length_num)
median(df_c_small$ride_length_num)
df_c_small$ride_length_min <- df_c_small$ride_length_num/60
skim(df_c_small)

ggplot(df_c_small, aes(ride_length_min)) +
  geom_histogram(binwidth = 0.5, col = "black", fill = "cornflowerblue") +
  labs(title="Frequency of ride lengths for casual riders",
       subtitle="From 10th to 90th quantile",
       x="ride length in minutes")
```

* __Interpretation:__ I cut off extreme values on both ends (10% each), so we only see the main distributions here. Annual members seem to make shorter rides than casual riders, most under 10 minutes. Casual riders' ride length peaks round 10 min, and there are also many rides that go up to 50 min. For annual members, this is rare. Casual riders maybe have a higher barrier to use the bikes, so they might rather use them for rides when it is 'worth it'.

```{r bar plot comparison weekdays by user category}
# bar plot for ride lengths per weekday (ordered), member groups next to each other
## make summary df for ggplot
df_weekday_member <- aggregate(df, by=list(df$day_of_week, df$member_casual), FUN=length)
df_weekday_member <- df_weekday_member %>% select("Group.1", "Group.2", "ride_id")
df_weekday_member <- df_weekday_member %>% rename(weekday = Group.1, member_casual = Group.2, count = ride_id)
#View(df_weekday_member)
df_weekday_member$weekday <- as.factor(df_weekday_member$weekday)
df_weekday_member$weekday <- ordered(df_weekday_member$weekday,
                                     levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday","Sunday"))

glimpse(df_weekday_member)
df_weekday_member$weekday 
## make bar chart
ggplot(df_weekday_member, aes(x=weekday,y=count,fill=member_casual)) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_discrete(name="Membership type",
                      breaks=c("casual","member"),
                      labels=c("Casual", "Member")) +
  xlab("Weekday") + ylab("Count") +
  scale_y_continuous(labels = comma) +
  labs(title="Rides per weekday and membership type")
# casual riders tend to ride more on weekends, members more during weekdays
```

* __Interpretation:__ Casual riders tend to ride more during weekends, annual members more during weekdays. This could mean that casual riders use the bikes in their freetime, annual members for work.

### Ideas to improve data / analysis
* There is no member id in the data (probably privacy reasons), but so we don't know how many rides members and casual riders make. If the company could provide this data safely, it could give more interesting insights.
* I could have spent way more time reporting the numbers, most of my interpretations rely only on the visualizations. This was a time-issue, and I might come back to this analysis later to give a more complete analysis and interpretation.
